Quantitative, Multispecies Monitoring at a Continental Scale

### 2025-04-07 Manuscript_plain_v3.tex
update

\section*{Author Contributions}
AOS, KMN, RPK, MS, and KP designed the research;\\
MRS, AR-L and AW did the laboratory work; \\
PB, RPK, and GG did the bioinformatics;\\
GG, OL, RPK, and AOS analyzed the data;\\
GG, OL, RPK, and AOS interpreted the results;\\
All authors provided edits to the manuscript \\

@@

\subsection*{MetabarcodFig. S3ing: Environmental Samples and Mock Community}
In total, 554 environmental samples were randomized and divided into 7 sequencing libraries for an Illumina MiSeq (v3 600 cycle kit). Total PCR reactions including PCR blanks (n = 7) and positive controls (n = 7), were amplified using MiFish-U universal primers \cite{miya2015} (see \href{SI_Appendix.pdf}{SI Appendix Extended Methods} for details on metabarcoding methods). Importantly, we only analyzed one water sample collected from each sampling station using metabarcoding.

@@

\subsection*{Statistical analysis}
To generate estimates of DNA concentration, we constructed a joint statistical model for qPCR and Mifish metabarcoding observations from the same samples. This model is based on the QM R package (www.github.com/gledguri/QM), with some modifications and advancements. The full statistical details are described in \href{SI_Appendix.pdf}{SI Appendix Extended Methods}, with an overview of the approach (Fig. \ref{fig:DAG_v3}). We used qPCR observations of hake DNA concentration for all water samples available (1,794 unique water samples were assayed and using a total of 5,394 qPCR reactions). This model takes observations of the PCR cycle at which amplification was detected in each field sample and estimates a hake DNA concentration for each water bottle collected at each sampling station-depth combination (Fig. \ref{fig:DAG_v3}). This statistical model accounts for sample-specific effects (e.g., dilution of the sample before PCR amplification), the effect of replicate water samples (a random effect), to provide estimates of hake DNA concentration in from each water sample collected at each location and depth (Fig. \ref{fig:DAG_v3}). This qPCR component of the joint model provides a quantification of DNA concentration for one species (Pacific hake) that can be connected to the multispecies proportions provided by DNA metabarcoding (Fig. \ref{fig:DAG_v3}).

\begin{figure*}[tbhp] 
\centering
\includegraphics[width=16cm]{plots/DAG_v3.pdf} 
\caption{Schematic for the statistical models to produce species distributions. Three sets of observations (orange) were used simultaneously to provide estimates of eDNA concentration for twelve fish species (blue) from which we derive smoothed predictions of DNA concentration (green). Mock communities for metabarcoding (yellow) were estimated externally and provided estimated $\alpha_s$ that were used in the joint model. Notation for parameters, subscripts, and observations are indicated in gray.}
\label{fig:DAG_v3}
\end{figure*}

@@

The above statistical model (Fig. \ref{fig:DAG_v3}) provides predictions of species-specific DNA concentrations at each location and depth sampled ($d$ = 0, 50, 150, 300, or 500 m). We used the sdmTMB package \cite{anderson2022} to generate smooth maps of DNA concentration for each species independently (Fig. \ref{fig:DAG_v3}, green box). As Pacific hake (\textit{Merluccius productus}) has been examined elsewhere using qPCR \cite{shelton2022}, we did not include the spatial patterns of that species in our analysis.

@@

Previous eDNA work has tended to feature more qualitative multispecies results, or else quantitative analysis of single species at small spatial scales. Here, we create a rigorous quantitative framework and use hundreds of water samples to make ecological inferences at the scale of management. In this study, we find spatial patterns of DNA concentrations reflect the distributions of commercially important coastal-pelagic fish species in three dimensions and across ten degrees of latitude in a single year of sampling in 2019 (Fig. \ref{fig:fig1}: Northern Anchovy, Pacific Sardine; \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}, Pacific Herring and Jack Mackerel). For other high-biomass species otherwise lacking well-described distributions, we provide the first species distribution models (Fig. \ref{fig:fig1}, \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}). We find that the estimated eDNA concentrations are correlated among species with similar traits and habitat preferences (Fig. \ref{fig:fig2}). Emergent patterns of species richness and estimated eDNA aggregation correspond with known ecological drivers, evidence that eDNA traces can reliably capture ecologically meaningful patterns (Fig. \ref{fig:fig3}). Particularly for forage fish species -- which are critical high-biomass links in marine food webs, but difficult to survey with traditional methods \cite{Pikitch2012} -- molecular methods offer atractable and quantitative index of abundance over large spatial scales.

Below we highlight results for six species (Fig. \ref{fig:fig1}) representing a range of habitat associations and general species information; distribution maps for all the remaining species are presented in \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}. In particular, our analyses present detailed, depth-specific information on abundance that is difficult to obtain with traditional sampling methods.

\begin{figure*}[tbhp]
\centering
\includegraphics[width=16cm]{plots/1_Figure_1.jpg} 
\caption{Estimated species (n=6; for the remaining species see \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}) eDNA concentration (A) across 0, 50, 150, 300, and 500m depth samples and known depth distribution of those species from literature (B).}
\label{fig:fig1}
\end{figure*}

\subsection*{Surface species}
Northern anchovy (\textit{Engraulis mordax}) and Pacific sardine (\textit{Sardinops sagax}) are the two iconic coastal pelagic species of the eastern North Pacfic \cite{steinbeck1945} and play a vital ecological role linking lower and upper trophic levels (e.g., Chinook salmon diets, sea lions, \cite{kaplan2019}); historically these supported large commercial fisheries particularly in California. While larvae and juveniles typically occupy the upper 50m, adults move to 100m during the day to avoid predators. Our eDNA estimates consistently reflect the expected three-dimensional distributions of known surface dwelling species, with high concentrations in surface waters for both species\cite{litz2008,parnel2008}, and an accumulation pocket near the Columbia River outflow (46.2$^\circ$N; Fig. \ref{fig:fig1}). Sardine favors slightly warmer temperatures and more southerly waters than anchovy (SST range from literature: 13–24$^\circ$C for sardine vs. 10–14$^\circ$C for anchovy). eDNA patterns tracked these differences with sardine eDNA in greatest abundance at 41$^\circ$N (range 40$^\circ$–45$^\circ$), while anchovy eDNA abundance was greatest at 43.5$^\circ$N (range 42$^\circ$–48$^\circ$; Fig. \ref{fig:fig4}) such that eDNA captured subtle yet meaningful differences in habitat preferences. When comparing the magnitude difference between the two species, eDNA abundance patterns suggest a 2.5-fold higher abundance for anchovy than sardine which agree in direction with stock assessments for anchovy and sardine (an approximately 35-fold difference; \cite{kuriyama2022, kuriyama2022b}). However, due to differences in stock definitions and survey areas (the anchovy stock assessment includes southern California waters (32.5$^\circ$–36$^\circ$N) not covered by the eDNA survey) these magnitudes are not directly comparable.

\subsection*{Midwater species}
Widow rockfish (\textit{Sebastes entomelas}) is common between 100-350m between northern Baja California and southern Alaska, with pelagic larvae and juveniles generally detected in shallower habitats ($<$ 150m depth) \cite{reynolds2001,bosley2014} including in nearshore kelp forests \cite{quigley2024, laidig2007} . Moreover, because it is a common bycatch species in the Pacific hake (\textit{Merluccius productus}) fishery, concerns about bycatch of widow rockfish can also influence the much larger hake fishery (i.e., it is a “choke” species for hake; \cite{somers2018}). Despite its commercial importance, no targeted survey effort exists for widow rockfish, in part because bottom trawls do not sample its pelagic habitat well. We find widow rockfish eDNA strongly and nearly exclusively associated with 50m sample depths (Fig. \ref{fig:fig1} and \ref{fig:fig4}), consistent with the pelagic larvae and juveniles reported in the literature and somewhat shallower than adults \cite{reynolds2001,bosley2014,quigley2024}, potentially reflecting the species’ nocturnal feeding schooling behavior \cite{reynolds2001,bosley2014}.

@@

\subsection*{Deepwater species}
Among deeper-water fishes, eDNA revealed spatial and potentially ontogenetic migration patterns for species that are common, yet under studied, such as the dragonfish (\textit{Tactostoma macropus}). Available literature suggests dragonfish occur roughly between 40$^\circ$N-50$^\circ$N (adults at 300-900m, larvae at 0-60m in summer) \cite{kawaguchi1993} and perform diel vertical migrations to around 100 meters \cite{kawaguchi1993,willis1982}. The eDNA concentrations appeared to reflect this activity (Fig. \ref{fig:fig4}), with the predominant surface signal at northern latitudes likely reflecting larvae and juveniles (Fig. \ref{fig:fig1}; near Columbia River discharge), and two deeper hotspots at 300m and 500m separated by about eight degrees of latitude. However, no other quantitative estimates of abundance or detailed distribution patterns for dragonfish exist in the current literature.

Also consistent with known habitats are commercially caught flatfish such as Dover sole (\textit{Microstomus pacificus}), which lives at 200m to 1200m along the Pacific coast \cite{drazen2007,brodziak2000,drazen2012}; individuals move progressively deeper as they grow \cite{vetter1994,hunter1990}. The eDNA concentration estimates are consistent with this pattern, showing high concentrations mostly in water samples collected close to the seabed (Fig. \ref{fig:fig1}) \cite{ono2016}.

@@

Pairwise correlations of DNA concentration showed meaningful patterns among species associated with depth (Fig. \ref{fig:fig2}). Surface-associated species were nearly all positively correlated with each other, with some correlations particularly large (Pearson's $\rho > 0.50$; Fig. \ref{fig:fig2}). Correlations were weak among the six midwater and deepwater species (all $\rho < |0.40|$; Fig. \ref{fig:fig2}). Within the surface species, further subdivision was also apparent, with chub mackerel (\textit{Scomber japonicus}) and sardine (\textit{Sardinops sagax}) showing very high correlation ($\rho = 0.76$; Fig. \ref{fig:fig2}) with both more common in the more southern (and somewhat warmer) waters (Fig. \ref{fig:fig1} and Fig. \ref{fig:fig4}). Concentrations of the remaining surface-dwelling species (Northern anchovy, eulachon, jack mackerel, and Pacific herring) had modest correlations with midwater species (Fig. \ref{fig:fig2}) consistent with more northerly distribution (Fig. \ref{fig:fig4}). Together these findings suggest that eDNA is capturing both depth and latitudinal habitat associations. Deep-water species are distinct (Fig. \ref{fig:fig1} and Fig. \ref{fig:fig4}) and not strongly correlated with those from other zones (Fig. \ref{fig:fig2}). Deeper water tends to feature lower and more diffuse fish eDNA concentrations (Fig. \ref{fig:fig1} and Fig. \ref{fig:fig4}). Lower overall eDNA concentrations and fewer deep water samples may make similarities among species distributions more difficult to detect at 300m and deeper.

\begin{figure}[tbhp] 
\centering
\includegraphics[width=10cm]{plots/2_Figure_2.jpg} 
\caption{Correlation plot describing species co-abundance and co-occurance. Spearman’s $\rho$ correlation coefficient are expressed in numbers and colors ranging from -1 (red) negatively correlated to 1 (blue) positively correlated. Species are ordered in known ecological groups (surface, midwater, and deepwater) to visually indicate the gradual change of species co-occurance.}
\label{fig:fig2}
\end{figure}

\subsection*{Species richness and total eDNA accumulation}
In addition to providing species-specific patterns, eDNA metrics of abundance from individual species can be combined to reveal important aggregate ecological patterns. For the 12 species considered here, we show how species richness (see \href{SI_Appendix.pdf}{SI Appendix, Fig. S2} for defining species presence) from eDNA varied substantially and predictably with depth, with the highest richness observed at the surface (Fig. \ref{fig:fig3}\textit{A}). In part, this pattern is likely to be an artifact of the focal species included in this study -- chosen for their relevance to fisheries management as well as for their frequency in the observations (see Methods) -- but the richness-depth gradient also reflects the expectation that habitat and nutrient availability is greatest near the surface \cite{smith2002,hickey2005}. Surface waters also host the eggs, larvae, and juvenile stages of many meso- and bathypelagic species \cite{parnel2008}, some of which we have included here, and hence we both expect and observe higher richness in the upper layer of the water column \cite{kim2011}.

Species richness also corresponds strongly with the influence of Columbia River plume (Fig. \ref{fig:fig3}\textit{A}), which extends well beyond the coastline \cite{hickey2003}. The nitrate-rich river water interacts with ocean currents and intensifies coastal upwelling, enhancing both primary and secondary productivity \cite{hickey2010} and likely influencing the patterns of species richness we observe (Fig. \ref{fig:fig3}\textit{A}) \cite{tolimieri2015}. The total fish eDNA accumulation (Fig. \ref{fig:fig3}\textit{B}) suggests the highest aggregation of forage fish DNA near Heceta Bank (44 degrees N; overlapping anchovy and sardine distributions are the two largest contributors to the observed peak of accumulated eDNA concentration) in addition to other commercialy important mesopelagic fish (\textit{Sebastes entomelas} and \textit{Tarletonbeania crenularis}) a center of abundance for commercial fisheries \cite{bosley2014,Tissot2008TwentyYO}, and elsewhere nearshore in areas of coastal upwelling.

\begin{figure*}[tbhp] 
\centering
\includegraphics[width=13cm]{plots/3_Figure_3.jpg} 
\caption{Spatial and vertical distribution of species richness (A) and environmental DNA (B) concentration across depths (0, 50, 150, 300, and 500 m). (B) Species richness represented as the cumulative number of species present (species with higher than 1.5 copies/L (see Fig. \ref{fig:fig4} for defining presence) on smoothed estimated eDNA concentrations) and acumulated eDNA concentration representing the cumulative sum of eDNA (copies/L) across all species.}
\label{fig:fig3}
\end{figure*}

\begin{figure*}[tbhp] 
\centering
\includegraphics[width=16cm]{plots/4_Figure_4.jpg} 
\caption{Vertical (A) and latitudinal (B) distribution of eDNA concentrations grouped in ecological groups (surface south, surface north, midwater, and deepwater). Mean eDNA concentration of vertical distribution is shown with dots and 99\% upper quantile with ticks (A). Probability density functions (PDF) of estimated eDNA concentrations across latitudinal gradients (B; square root transformed of x-axis to enhance low concentration peaks). Key geographic features, such as the Columbia River and Heceta Bank, are marked to indicate alignment of species distributions.}
\label{fig:fig4}
\end{figure*}

@@

The majority of eDNA shed into the environment dissipates from surface waters within hours \cite{andruszkiewiczallan2021}. Thus, the horizontal scale of environmental smoothing in our dataset likely spans several kilometers at most \cite{baetscher2023}. Given that our sampling area spans more than 1,200 km north to south and tens of km east to west, the spatial patterns we report are likely to reflect genuine differences in fish communities rather than artifacts of eDNA transport dynamics \cite{guri2023}. Vertical smoothing of eDNA signals appears minimal, with patterns at the scale of meters: shallow-water species (sardine, anchovy, herring) occur largely at the surface and 50m depth bins; Widow rockfish occurs almost exclusively at the 50m depth; species found throughout the water column are either diel vertical migrators or species occupying different depths at various life-history stages (Fig. \ref{fig:fig3}\textit{B}); and bottom-dwelling fishes are detected primarily at the deepest point sampled on any station.

Notably, the comparison between sample-specific eDNA estimation (\href{SI_Appendix.pdf}{SI Appendix, Fig. S6}) and spatially smoothed maps (Fig. \ref{fig:fig1} and \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}) revealed minimal differences. Large-scale sampling combined with statistical smoothing mitigates many challenges associated with eDNA transport and degradation -- which decouple the eDNA observations from their particular source organisms -- allowing reliable detection of broad patterns in species distributions. Further methodological improvements may incorporate information about the ecology of eDNA itself, such as the degree to which observations of eDNA concentrations depend upon water temperature or other covariates \cite{jo2019b}.

>>>>

\section*{Author Contributions}
AOS, KMN, RPK, MS, and KP designed the research;
MRS, AR-L and AW did the laboratory work; 
PB, RPK, and GG did the bioinformatics;
GG, OL, RPK, and AOS analyzed the data;
GG, OL, RPK, and AOS interpreted the results;
All authors provided edits to the manuscript \\

@@

\subsection*{Metabarcoding: Environmental Samples and Mock Community}
In total, 554 environmental samples were randomized and divided into 7 sequencing libraries for an Illumina MiSeq (v3 600 cycle kit). Total PCR reactions including PCR blanks (n = 7) and positive controls (n = 7), were amplified using MiFish-U universal primers \cite{miya2015} (see \href{SI_Appendix.pdf}{SI Appendix Extended Methods} for details on metabarcoding methods). Importantly, we only analyzed one water sample collected from each sampling station using metabarcoding.

@@

\subsection*{Statistical analysis}
To generate estimates of DNA concentration, we constructed a joint statistical model for qPCR and Mifish metabarcoding observations from the same samples. This model is based on the QM R package (www.github.com/gledguri/QM), with some modifications and advancements. The full statistical details are described in \href{SI_Appendix.pdf}{SI Appendix Extended Methods}, with an overview of the approach (Fig. \ref{fig:fig1}). We used qPCR observations of hake DNA concentration for all water samples available (1,794 unique water samples were assayed and using a total of 5,394 qPCR reactions). This model takes observations of the PCR cycle at which amplification was detected in each field sample and estimates a hake DNA concentration for each water bottle collected at each sampling station-depth combination (Fig. \ref{fig:fig1}). This statistical model accounts for sample-specific effects (e.g., dilution of the sample before PCR amplification), the effect of replicate water samples (a random effect), to provide estimates of hake DNA concentration in from each water sample collected at each location and depth (Fig. \ref{fig:fig1}). This qPCR component of the joint model provides a quantification of DNA concentration for one species (Pacific hake) that can be connected to the multispecies proportions provided by DNA metabarcoding (Fig. \ref{fig:fig1}).

\begin{figure}[tbhp] 
\centering
\includegraphics[width=16cm]{plots/DAG_v3.pdf} 
\caption{Schematic for the statistical models to produce species distributions. Three sets of observations (orange) were used simultaneously to provide estimates of eDNA concentration for twelve fish species (blue) from which we derive smoothed predictions of DNA concentration (green). Mock communities for metabarcoding (yellow) were estimated externally and provided estimated $\alpha_s$ that were used in the joint model. Notation for parameters, subscripts, and observations are indicated in gray.}
\label{fig:fig1}
\end{figure}

@@

The above statistical model (Fig. \ref{fig:fig1}) provides predictions of species-specific DNA concentrations at each location and depth sampled ($d$ = 0, 50, 150, 300, or 500 m). We used the sdmTMB package \cite{anderson2022} to generate smooth maps of DNA concentration for each species independently (Fig. \ref{fig:fig1}, green box). As Pacific hake (\textit{Merluccius productus}) has been examined elsewhere using qPCR \cite{shelton2022}, we did not include the spatial patterns of that species in our analysis.

@@

Previous eDNA work has tended to feature more qualitative multispecies results, or else quantitative analysis of single species at small spatial scales. Here, we create a rigorous quantitative framework and use hundreds of water samples to make ecological inferences at the scale of management. In this study, we find spatial patterns of DNA concentrations reflect the distributions of commercially important coastal-pelagic fish species in three dimensions and across ten degrees of latitude in a single year of sampling in 2019 (Fig. \ref{fig:fig2}: Northern Anchovy, Pacific Sardine; \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}, Pacific Herring and Jack Mackerel). For other high-biomass species otherwise lacking well-described distributions, we provide the first species distribution models (Fig. \ref{fig:fig2}, \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}). We find that the estimated eDNA concentrations are correlated among species with similar traits and habitat preferences (Fig. \ref{fig:fig3}). Emergent patterns of species richness and estimated eDNA aggregation correspond with known ecological drivers, evidence that eDNA traces can reliably capture ecologically meaningful patterns (Fig. \ref{fig:fig4}). Particularly for forage fish species -- which are critical high-biomass links in marine food webs, but difficult to survey with traditional methods \cite{Pikitch2012} -- molecular methods offer atractable and quantitative index of abundance over large spatial scales.

Below we highlight results for six species (Fig. \ref{fig:fig2}) representing a range of habitat associations and general species information; distribution maps for all the remaining species are presented in \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}. In particular, our analyses present detailed, depth-specific information on abundance that is difficult to obtain with traditional sampling methods.

\begin{figure}[tbhp]
\centering
\includegraphics[width=0.9\textwidth, height=1.17\textwidth]{plots/1_Figure_1.pdf} 
\caption{Estimated species (n=6; for the remaining species see \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}) eDNA concentration (A) across 0, 50, 150, 300, and 500m depth samples and known depth distribution of those species from literature (B).}
\label{fig:fig2}
\end{figure}

\subsection*{Surface species}
Northern anchovy (\textit{Engraulis mordax}) and Pacific sardine (\textit{Sardinops sagax}) are the two iconic coastal pelagic species of the eastern North Pacfic \cite{steinbeck1945} and play a vital ecological role linking lower and upper trophic levels (e.g., Chinook salmon diets, sea lions, \cite{kaplan2019}); historically these supported large commercial fisheries particularly in California. While larvae and juveniles typically occupy the upper 50m, adults move to 100m during the day to avoid predators. Our eDNA estimates consistently reflect the expected three-dimensional distributions of known surface dwelling species, with high concentrations in surface waters for both species\cite{litz2008,parnel2008}, and an accumulation pocket near the Columbia River outflow (46.2$^\circ$N; Fig. \ref{fig:fig2}). Sardine favors slightly warmer temperatures and more southerly waters than anchovy (SST range from literature: 13–24$^\circ$C for sardine vs. 10–14$^\circ$C for anchovy). eDNA patterns tracked these differences with sardine eDNA in greatest abundance at 41$^\circ$N (range 40$^\circ$–45$^\circ$), while anchovy eDNA abundance was greatest at 43.5$^\circ$N (range 42$^\circ$–48$^\circ$; Fig. \ref{fig:fig5}) such that eDNA captured subtle yet meaningful differences in habitat preferences. When comparing the magnitude difference between the two species, eDNA abundance patterns suggest a 2.5-fold higher abundance for anchovy than sardine which agree in direction with stock assessments for anchovy and sardine (an approximately 35-fold difference; \cite{kuriyama2022, kuriyama2022b}). However, due to differences in stock definitions and survey areas (the anchovy stock assessment includes southern California waters (32.5$^\circ$–36$^\circ$N) not covered by the eDNA survey) these magnitudes are not directly comparable.

\subsection*{Midwater species}
Widow rockfish (\textit{Sebastes entomelas}) is common between 100-350m between northern Baja California and southern Alaska, with pelagic larvae and juveniles generally detected in shallower habitats ($<$ 150m depth) \cite{reynolds2001,bosley2014} including in nearshore kelp forests \cite{quigley2024, laidig2007} . Moreover, because it is a common bycatch species in the Pacific hake (\textit{Merluccius productus}) fishery, concerns about bycatch of widow rockfish can also influence the much larger hake fishery (i.e., it is a “choke” species for hake; \cite{somers2018}). Despite its commercial importance, no targeted survey effort exists for widow rockfish, in part because bottom trawls do not sample its pelagic habitat well. We find widow rockfish eDNA strongly and nearly exclusively associated with 50m sample depths (Fig. \ref{fig:fig2} and \ref{fig:fig5}), consistent with the pelagic larvae and juveniles reported in the literature and somewhat shallower than adults \cite{reynolds2001,bosley2014,quigley2024}, potentially reflecting the species’ nocturnal feeding schooling behavior \cite{reynolds2001,bosley2014}.

@@

\subsection*{Deepwater species}
Among deeper-water fishes, eDNA revealed spatial and potentially ontogenetic migration patterns for species that are common, yet under studied, such as the dragonfish (\textit{Tactostoma macropus}). Available literature suggests dragonfish occur roughly between 40$^\circ$N-50$^\circ$N (adults at 300-900m, larvae at 0-60m in summer) \cite{kawaguchi1993} and perform diel vertical migrations to around 100 meters \cite{kawaguchi1993,willis1982}. The eDNA concentrations appeared to reflect this activity (Fig. \ref{fig:fig5}), with the predominant surface signal at northern latitudes likely reflecting larvae and juveniles (Fig. \ref{fig:fig2}; near Columbia River discharge), and two deeper hotspots at 300m and 500m separated by about eight degrees of latitude. However, no other quantitative estimates of abundance or detailed distribution patterns for dragonfish exist in the current literature.

Also consistent with known habitats are commercially caught flatfish such as Dover sole (\textit{Microstomus pacificus}), which lives at 200m to 1200m along the Pacific coast \cite{drazen2007,brodziak2000,drazen2012}; individuals move progressively deeper as they grow \cite{vetter1994,hunter1990}. The eDNA concentration estimates are consistent with this pattern, showing high concentrations mostly in water samples collected close to the seabed (Fig. \ref{fig:fig2}) \cite{ono2016}.

@@

Pairwise correlations of DNA concentration showed meaningful patterns among species associated with depth (Fig. \ref{fig:fig3}). Surface-associated species were nearly all positively correlated with each other, with some correlations particularly large (Pearson's $\rho > 0.50$; Fig. \ref{fig:fig3}). Correlations were weak among the six midwater and deepwater species (all $\rho < |0.40|$; Fig. \ref{fig:fig3}). Within the surface species, further subdivision was also apparent, with chub mackerel (\textit{Scomber japonicus}) and sardine (\textit{Sardinops sagax}) showing very high correlation ($\rho = 0.76$; Fig. \ref{fig:fig3}) with both more common in the more southern (and somewhat warmer) waters (Fig. \ref{fig:fig2} and Fig. \ref{fig:fig5}). Concentrations of the remaining surface-dwelling species (Northern anchovy, eulachon, jack mackerel, and Pacific herring) had modest correlations with midwater species (Fig. \ref{fig:fig3}) consistent with more northerly distribution (Fig. \ref{fig:fig5}). Together these findings suggest that eDNA is capturing both depth and latitudinal habitat associations. Deep-water species are distinct (Fig. \ref{fig:fig2} and Fig. \ref{fig:fig5}) and not strongly correlated with those from other zones (Fig. \ref{fig:fig3}). Deeper water tends to feature lower and more diffuse fish eDNA concentrations (Fig. \ref{fig:fig2} and Fig. \ref{fig:fig5}). Lower overall eDNA concentrations and fewer deep water samples may make similarities among species distributions more difficult to detect at 300m and deeper.

\begin{figure}[tbhp]
\centering
%\includegraphics[width=10cm, height = 13cm]{plots/2_Figure_2.jpg} 
\includegraphics[width=0.6\textwidth, height=0.6\textwidth]{plots/2_Figure_2.pdf} 
\caption{Correlation plot describing species co-abundance and co-occurance. Spearman’s $\rho$ correlation coefficient are expressed in numbers and colors ranging from -1 (red) negatively correlated to 1 (blue) positively correlated. Species are ordered in known ecological groups (surface, midwater, and deepwater) to visually indicate the gradual change of species co-occurance.}
\label{fig:fig3}
\end{figure}

\subsection*{Species richness and total eDNA accumulation}
In addition to providing species-specific patterns, eDNA metrics of abundance from individual species can be combined to reveal important aggregate ecological patterns. For the 12 species considered here, we show how species richness (see \href{SI_Appendix.pdf}{SI Appendix, Fig. S2} for defining species presence) from eDNA varied substantially and predictably with depth, with the highest richness observed at the surface (Fig. \ref{fig:fig4}\textit{A}). In part, this pattern is likely to be an artifact of the focal species included in this study -- chosen for their relevance to fisheries management as well as for their frequency in the observations (see Methods) -- but the richness-depth gradient also reflects the expectation that habitat and nutrient availability is greatest near the surface \cite{smith2002,hickey2005}. Surface waters also host the eggs, larvae, and juvenile stages of many meso- and bathypelagic species \cite{parnel2008}, some of which we have included here, and hence we both expect and observe higher richness in the upper layer of the water column \cite{kim2011}.

Species richness also corresponds strongly with the influence of Columbia River plume (Fig. \ref{fig:fig4}\textit{A}), which extends well beyond the coastline \cite{hickey2003}. The nitrate-rich river water interacts with ocean currents and intensifies coastal upwelling, enhancing both primary and secondary productivity \cite{hickey2010} and likely influencing the patterns of species richness we observe (Fig. \ref{fig:fig4}\textit{A}) \cite{tolimieri2015}. The total fish eDNA accumulation (Fig. \ref{fig:fig4}\textit{B}) suggests the highest aggregation of forage fish DNA near Heceta Bank (44 degrees N; overlapping anchovy and sardine distributions are the two largest contributors to the observed peak of accumulated eDNA concentration) in addition to other commercialy important mesopelagic fish (\textit{Sebastes entomelas} and \textit{Tarletonbeania crenularis}) a center of abundance for commercial fisheries \cite{bosley2014,Tissot2008TwentyYO}, and elsewhere nearshore in areas of coastal upwelling.

\begin{figure}[tbhp] 
\centering
\includegraphics[width=13cm]{plots/3_Figure_3.pdf} 
\caption{Spatial and vertical distribution of species richness (A) and environmental DNA (B) concentration across depths (0, 50, 150, 300, and 500 m). (B) Species richness represented as the cumulative number of species present (species with higher than 1.5 copies/L (see \href{SI_Appendix.pdf}{SI Appendix, Fig. S2} for defining presence) on smoothed estimated eDNA concentrations) and acumulated eDNA concentration representing the cumulative sum of eDNA (copies/L) across all species.}
\label{fig:fig4}
\end{figure}

\begin{figure}[tbhp] 
\centering
\includegraphics[width=16cm]{plots/4_Figure_4.pdf} 
\caption{Vertical (A) and latitudinal (B) distribution of eDNA concentrations grouped in ecological groups (surface south, surface north, midwater, and deepwater). Mean eDNA concentration of vertical distribution is shown with dots and 99\% upper quantile with ticks (A). Probability density functions (PDF) of estimated eDNA concentrations across latitudinal gradients (B; square root transformed of x-axis to enhance low concentration peaks). Key geographic features, such as the Columbia River and Heceta Bank, are marked to indicate alignment of species distributions.}
\label{fig:fig5}
\end{figure}

@@

The majority of eDNA shed into the environment dissipates from surface waters within hours \cite{andruszkiewiczallan2021}. Thus, the horizontal scale of environmental smoothing in our dataset likely spans several kilometers at most \cite{baetscher2023}. Given that our sampling area spans more than 1,200 km north to south and tens of km east to west, the spatial patterns we report are likely to reflect genuine differences in fish communities rather than artifacts of eDNA transport dynamics \cite{guri2023}. Vertical smoothing of eDNA signals appears minimal, with patterns at the scale of meters: shallow-water species (sardine, anchovy, herring) occur largely at the surface and 50m depth bins; Widow rockfish occurs almost exclusively at the 50m depth; species found throughout the water column are either diel vertical migrators or species occupying different depths at various life-history stages (Fig. \ref{fig:fig4}\textit{B}); and bottom-dwelling fishes are detected primarily at the deepest point sampled on any station.

Notably, the comparison between sample-specific eDNA estimation (\href{SI_Appendix.pdf}{SI Appendix, Fig. S6}) and spatially smoothed maps (Fig. \ref{fig:fig2} and \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}) revealed minimal differences. Large-scale sampling combined with statistical smoothing mitigates many challenges associated with eDNA transport and degradation -- which decouple the eDNA observations from their particular source organisms -- allowing reliable detection of broad patterns in species distributions. Further methodological improvements may incorporate information about the ecology of eDNA itself, such as the degree to which observations of eDNA concentrations depend upon water temperature or other covariates \cite{jo2019b}.
